Union and intersection contracts are hard, actually

### 2021-06-10 paper.tex
Move and rework example of union

In the example above, the first term prints \nickel{"hi"} two times while the
second term prints it only once. However, \ref{eq:cbn-expansion} does hold for
\emph{pure} terms, that are terms without side-effects. In a pure language, or a
language with effect tracking like Haskell\todo{Cite a different language, or
remove citation}, pure terms can be safely identified for applying
optimizations.

Strikingly, we will see in section~\ref{sec:issues-sem} that the introduction of
unions and intersections can make \ref{eq:cbn-expansion} unsound
\emph{even in a pure setting}, making the optimization of
Figure~\ref{fig:optimizations-let-floating-ex} unsound in general.

@@

Like for static types, value which satisfies the contract \nickel{A @| B}
is a value which satisfy either contract \nickel{A} or
contract \nickel{B} (though in Section~\ref{sec:issues-literature} we
will see that it may be desirable to weaken this definition).

@@

\resolved{Arnaud: this paragraph is pretty good, on the other hand, there
is no value which lets me complete the example, so this example falls
a bit flat.
Teo: Improved the example, also added another one commented that is, for me,
more realistic, but I think the one uncommented is simpler.}
We have shown how values of a union are built, is worth mentioning
how they are used.
For instance, consider the TypeScript function on Figure \ref{fig:addElem},
we want it to insert an element to a list that is either an
\typescript{Array< 42 | "Nickel" >} or an \typescript{Array< "Nickel" | 3.5 >},
where \typescript{42}, \typescript{3.5}, and \typescript{"Nickel"} are literal
types\footnote{Note that in TypeScript, union and intersection are written as
\typescript{|} and \typescript{&}, instead of \nickel{@|} and \nickel{@&},
as we do in Nickel.}.
What value could we fill the square ($\square$) with for this to typecheck?
Since we don't know what's the shape of \typescript{list},
we can't know whether to push a \typescript{42 | "Nickel"} or a
\typescript{"Nickel" | 3.5}. In general,
when using a union, we have to treat it as if it were any one
of the alternatives. In the case of the example, we should provide
it with a value that could be pushed to both arrays, the only possible one
is \typescript{"Nickel"},
since it's the only value that satisfies the intersection of
\typescript{42 | "Nickel"} and \typescript{"Nickel" | 3.5}.

\info{Not planned, but nice place to smooth things out towards
intersections}

\begin{figure}[h]
\begin{lstlisting}[language=JavaScript]
function addElem(list: Array< 42 | "Nickel" > | Array< "Nickel" | 3.5 >) {
    list.push((*@ $\square$ @*))
}
\end{lstlisting}
\caption{Adding an element to a union of two arrays}
\label{fig:addElem}
\end{figure}

@@

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let Animal =
  { species: Str, breed: Str, name: Str } in
let Pet = { owner: Str, name: Str } in
let myDog | Animal @& Pet =
  { species = "Canis Lupus",
    breed   = "Australian Cattle Dog",
    owner   = "Anonymous Author",
    name    = "Juno" }
\end{lstlisting}
\caption{An animal that is also a pet}
\label{fig:intersection-record}
\end{figure}

>>>>

In the example above, the first term prints \nickel{"hi"} two times while the
second term prints it only once. However, \ref{eq:cbn-expansion} does hold for
\emph{pure} terms, that are terms without side-effects. In a pure language, or a
language with effect tracking like PureScript\resolved{Cite a different
    language, or remove citation. (ADDED) tbh PureScript is still
statically-typed, compiled lang, but who knows: the "script" in the name and the
web orientation may still be more appealing to the audience of DLS}, pure terms
can be safely identified for applying optimizations.

Strikingly, we will see in section~\ref{sec:issues-sem} that the introduction of
unions and intersections can make \ref{eq:cbn-expansion} unsound \emph{even in a
pure setting}, preventing the optimization of
Figure~\ref{fig:optimizations-let-floating-ex} to fire in general.

@@

Like for static types, a value which satisfies the contract \nickel{A @| B}
is a value which satisfies either contract \nickel{A} or
contract \nickel{B} (though in Section~\ref{sec:issues-literature} we
will see that it may be desirable to weaken this definition).

@@

\resolved{Arnaud: this paragraph is pretty good, on the other hand, there
is no value which lets me complete the example, so this example falls
a bit flat.
Teo: Improved the example, also added another one commented that is, for me,
more realistic, but I think the one uncommented is simpler.}

@@

Figure~\ref{fig:addElem} gives a concrete example of this phenomenon. The
function \nickel{appendDate} append an element to \nickel{list} whose type is
only known to be in union of two lists, each using a different representation.
Both alternatives support falling back to a simple string for unparsed dates.
Faced with these two possibilities, \nickel{appendDate} can only append a
value which fits both types: this is precisely the intersection
\nickel{(Date @| Str) @& (DateWeek @| Str)}, that is, \nickel{Str}.

\begin{figure}[h]
\begin{lstlisting}[language=JavaScript]
let Date = {day | Num, month | Num, year | Num} in
let DateWeek = {dayOfWeek | Num, week | Num, year | Num} in
let appendDate | (List (Date @| Str)
                 @| List (DateWeek @| Str))
                 -> List
  = fun list =>
  lists.cons list "01/01/2021"
}
\end{lstlisting}
\caption{Adding an element to a union of two arrays}
\label{fig:addElem}
\end{figure}

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let Animal =
  { species | Str, breed | Str, name | Str } in
let Pet = { owner | Str, name | Str } in
let myDog | Animal @& Pet =
  { species = "Canis Lupus",
    breed   = "Australian Cattle Dog",
    owner   = "Anonymous Author",
    name    = "Juno" }
\end{lstlisting}
\caption{An animal that is also a pet}
\label{fig:intersection-record}
\end{figure}
